Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-02-02 paper/paper.tex
Conclusion and spacing

\begin{figure}[b]
    \centering
    \begin{tikzpicture}[scale=0.5]
        \begin{scope}
            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (Na) at (1,  0) {1};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (Nb) at (1, -1.5) {2};
            \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (Nc) at (0, -3) {3};
            \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (Nd) at (2, -3) {4};

@@

            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (N1) at (5.5,  0) {1};
            \node [draw, circle, fill=white, inner sep=1.5pt] (N2) at (7.5,  0) {2};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (N3) at (5.5, -1.5) {3};
            \node [draw, circle, fill=white, inner sep=1.5pt] (N4) at (7.5, -1.5) {4};
            \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (N5) at (5.5, -3) {5};
            \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (N6) at (7.5, -3) {6};

@@

            \node [draw, circle, fill=white, inner sep=1.5pt] (M1) at (-5.5,  0) {1};
            \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (M2) at (-3.5,  0) {2};
            \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (M3) at (-5.5, -1.5) {3};
            \node [draw, circle, fill=white, inner sep=1.5pt] (M4) at (-3.5, -1.5) {4};
            \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (M5) at (-5.5, -3) {5};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (M6) at (-3.5, -3) {6};

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
    \end{tikzpicture}\\[0.75cm]
    \begin{tikzpicture}[scale=0.5]
        \begin{scope}
            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (N1\n) at (\n, 0) {\n};
            }

@@

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (N2\n) at (0, -\n) {\n};
            }

@@

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (N3\n) at (7, -\n) {\n};
            }

@@

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (N4\n) at (\n, -7) {\n};
            }

@@

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices. Such
instances are guaranteed to be satisfiable---\citet{Anton:2009} exploited this property to create
large sets of random satisfiable boolean satisfiability instances.  However, since this has been the
only approach used to generate random subgraph isomorphism instances, existing benchmark suites
contain relatively few non-trivial unsatisfiable instances, and the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has lead to bias in algorithm design, to the extent that some proposed techniques will
\emph{only} work on satisfiable instances \citet{Battiti:2007}.

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and
only if
$v$ and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with
$n$ vertices, and an edge between each distinct pair of vertices with independent probability $p$.

@@

Our experiments were performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes RAM
(although this much RAM was not needed), running Scientific Linux release 6.7. We selected three
subgraph isomorphism solvers: the Glasgow solver \citep{McCreesh:2015}, LAD \citep{Solnon:2010}, and
VF2 \citep{Cordella:2004}; each was compiled using GCC 4.9. These solvers all use backtracking
search to build up an assignment of target vertices (values) to pattern vertices (variables), with
different kinds of inference and ordering heuristics.

@@

The light line plots mean search effort using the Glasgow solver: for sparse patterns, the problem
is trivial, for dense patterns proving unsatisfiability is not particularly difficult, and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure 1 of \citet{LeytonBrown:2014}. In particular,
satisfiable instances tend to be easier, but show greater variation than unsatisfiable instances. % , and
% there are exceptionally hard satisfiable instances \citep{Smith:1997}.

@@

For larger patterns, more complex behaviour emerges. If we are in the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-p_d$ and $t_d$ directions.  However, when inside the remaining two eighths of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved (and perhaps
\citeauthor{Walsh:1998}'s suggestion applies only in these conditions). In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour.

@@

We also implemented an equivalent SAT encoding with direct encoding of the cardinality constraints.
Using the Glucose solver, we saw very similar performance to the PB encoding. We also implemented a
similar integer program encoding; the Gurobi solver was only able to solve some of the trivial
satisfiable instances, and was almost never able to prove unsatisfiability within the time limit. We
do not plot the results for either of these solvers.

The \emph{association graph encoding} of a subgraph isomorphism problem is constructed by creating a
new graph with a vertex for each pair $(p, t)$ of vertices from the pattern and target graphs
respectively. There is an edge between vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$
to $t_1$ and $p_2$ to $t_2$ simultaneously is permitted, i.e.\ $p_1$ is adjacent to $p_2$ if and
only if $t_1$ is adjacent to $t_2$. A clique of size equal to the order of the pattern graph exists
in the association graph if and only if the problem is satisfiable \citep{Levi:1973}.  We illustrate
this in \cref{figure:sip}. We used this encoding with a version of the BBMC clique algorithm
\citep{SanSegundo:2011} modified to solve the decision problem. Again, the instances in the central
region remain hard, and additionally, some of the easy satisfiable instances become hard.

@@

?? We found hard instances, some where we'd expect, some in unexpected places. These unexpected
instances, by and large, remained hard under reductions. We looked at how existing non-induced
heuristics could be recovered from Sol, and how this gives us better heuristics for the induced
variant.

?? Consider dynamic heuristics.

?? Other random models and non-random.

?? Does not give hard instances for graph isomorphism.

>>>>

\begin{figure}[b]
    \centering
    \begin{tikzpicture}[scale=0.45,every node/.style={font=\footnotesize}]
        \begin{scope}
            \node [draw, circle, fill=uofgcobalt, inner sep=1pt] (Na) at (1,  0) {1};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1pt] (Nb) at (1, -1.5) {2};
            \node [draw, circle, fill=uofglawn, inner sep=1pt] (Nc) at (0, -3) {3};
            \node [draw, circle, fill=uofgthistle, inner sep=1pt] (Nd) at (2, -3) {4};

@@

            \node [draw, circle, fill=uofgcobalt, inner sep=1pt] (N1) at (5.5,  0) {1};
            \node [draw, circle, fill=white, inner sep=1pt] (N2) at (7.5,  0) {2};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1pt] (N3) at (5.5, -1.5) {3};
            \node [draw, circle, fill=white, inner sep=1pt] (N4) at (7.5, -1.5) {4};
            \node [draw, circle, fill=uofglawn, inner sep=1pt] (N5) at (5.5, -3) {5};
            \node [draw, circle, fill=uofgthistle, inner sep=1pt] (N6) at (7.5, -3) {6};

@@

            \node [draw, circle, fill=white, inner sep=1pt] (M1) at (-5.5,  0) {1};
            \node [draw, circle, fill=uofgcobalt, inner sep=1pt] (M2) at (-3.5,  0) {2};
            \node [draw, circle, fill=uofgthistle, inner sep=1pt] (M3) at (-5.5, -1.5) {3};
            \node [draw, circle, fill=white, inner sep=1pt] (M4) at (-3.5, -1.5) {4};
            \node [draw, circle, fill=uofglawn, inner sep=1pt] (M5) at (-5.5, -3) {5};
            \node [draw, circle, fill=uofgpumpkin, inner sep=1pt] (M6) at (-3.5, -3) {6};

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
    \end{tikzpicture}\\[0.25cm]
    \begin{tikzpicture}[scale=0.45,every node/.style={font=\footnotesize}]
        \begin{scope}
            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgcobalt, inner sep=1pt] (N1\n) at (\n, 0) {\n};
            }

@@

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgpumpkin, inner sep=1pt] (N2\n) at (0, -\n) {\n};
            }

@@

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofglawn, inner sep=1pt] (N3\n) at (7, -\n) {\n};
            }

@@

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgthistle, inner sep=1pt] (N4\n) at (\n, -7) {\n};
            }

@@

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded
degree, or mesh) graph and permuting the vertices. Such instances are guaranteed to be
satisfiable---\citet{Anton:2009} exploited this property to create large sets of random satisfiable
boolean satisfiability instances.  However, since this has been the only approach used to generate
random subgraph isomorphism instances, existing benchmark suites contain relatively few non-trivial
unsatisfiable instances, and the satisfiable instances tend to be computationally fairly easy, with
most of the difficulty being in dealing with the size of the model. This has lead to bias in
algorithm design, to the extent that some proposed techniques will \emph{only} work on satisfiable
instances \citep{Battiti:2007}.

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops.  The \emph{order} of a
graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the vertex set of a
graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the graph with the
same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and only if $v$
and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with $n$
vertices, and an edge between each distinct pair of vertices with independent probability $p$.

@@

Our experiments were performed on systems with Intel Xeon E5-4650 v2 CPUs, running Scientific Linux
release 6.7. We selected three subgraph isomorphism solvers: the Glasgow solver
\citep{McCreesh:2015}, LAD \citep{Solnon:2010}, and VF2 \citep{Cordella:2004}; each was compiled
using GCC 4.9. Each uses backtracking search to build up an assignment of target vertices (values)
to pattern vertices (variables), with different kinds of inference and ordering heuristics.

@@

The light line plots mean search effort using the Glasgow solver: for sparse patterns, the problem
is trivial, for dense patterns proving unsatisfiability is not particularly difficult, and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure 1 of \citet{LeytonBrown:2014}. In particular,
satisfiable instances tend to be easier, but show greater variation than unsatisfiable instances, and
there are exceptionally hard satisfiable instances \citep{Smith:1997}.

@@

For larger patterns, more complex behaviour emerges. If we are in the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-p_d$ and $t_d$ directions.  However, when inside the remaining two eighths of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved (and perhaps
\citeauthor{Walsh:1998}'s suggestion applies only in these conditions). In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour in these two eighths of the parameter space.

@@

We also implemented (but do not plot) an equivalent SAT encoding with direct encoding of the
cardinality constraints.  Using the Glucose solver, we saw very similar performance to the PB
encoding. We also implemented a similar integer program encoding; the Gurobi solver was only able to
solve some of the trivial satisfiable instances, and was almost never able to prove unsatisfiability
within the time limit.

The \emph{association graph encoding} of a subgraph isomorphism problem (illustrated in
\cref{figure:sip}) is constructed by creating a new graph with a vertex for each pair $(p, t)$ of
vertices from the pattern and target graphs respectively. There is an edge between vertex $(p_1,
t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$ to $t_1$ and $p_2$ to $t_2$ simultaneously is
permitted, i.e.\ $p_1$ is adjacent to $p_2$ if and only if $t_1$ is adjacent to $t_2$. A clique of
size equal to the order of the pattern graph exists in the association graph if and only if the
problem is satisfiable \citep{Levi:1973}. We used this encoding with a decision version of the BBMC
clique algorithm \citep{SanSegundo:2011}. Again, the instances in the central region remain hard,
and additionally, some of the easy satisfiable instances become hard.

@@

We have shown how to generate small but hard instances for the non-induced and induced subgraph
isomorphism problems, which will help offset the bias in existing datasets. For non-induced
isomorphisms, behaviour was as in many other hard problems, but for induced isomorphisms we
uncovered several interesting phenomena: there are hard instances far from a phase transition,
constrainedness predicts this, and existing general techniques for designing heuristics do not work
in certain portions of the parameter space.

In future work, we intend to repeat the experiments using other random models, including bounded and
regular degree. We will also look at dynamic heuristics, and switching pattern and target heuristics
independently. Finally, we intend to investigate whether variance can be calculated efficiently
enough to give better predictions for very sparse or dense pattern graphs.
